---
title: "Trach Study Clinical Data"
author: "HST.953"
date: "6/3/2019"
output: html_document
---

#### Important Data

1. Day 1 of ICU
2. Day 1 of Mechvent
3. Day 7 of Mechvent

#### Modeling

1. 3mo Mortality & 1yr Mortality ON Day 7 of Mechvent

All data use a 0.25day buffer period to account for any database lag, consistent with other validated MIMIC-III scripts.

### Utility Functions

#### rename

Renames columns, avoiding some.

```{r}
rename <- function(dat, avoid_names, prefix){
        for (name in colnames(dat)[!(colnames(dat) %in% avoid_names)]){
            colnames(dat)[which(colnames(dat) == name)] <- paste(prefix, name, sep = '')
        }
    return(dat)
}
```

#### outlier_na

If a value is out of range, replace it with `NA`

```{r}
outlier_na <- function(dat, nm){
  for(var in nm$variable){
    ## Create dictionary of columns
    dictionary <- colnames(dat)[grepl(var, colnames(dat))]
    for (lab in dictionary){
      ## If value is out of range, replace with NA
      dat[[lab]] <- ifelse((dat[[lab]] > nm[nm$variable==var, ]$max) | (dat[[lab]] < nm[nm$variable == var, ]$min), NA, dat[[lab]])
      #print(as.data.frame(summary(dat[,grepl(lab, colnames(dat))])))
    }
  }
  return(dat)
}
```

#### Outlier replace

If a variable is missing, replace it will value from literature.

```{r}
outlier_replace <- function(dat, normals){
    ## For each normal variable
    for (val in normals$variable){
        ## Find all columns with the variable
        dictionary <- colnames(dat[ ,colnames(dat)[grepl(val, colnames(dat))]])
        ## Iterate through...
        for (nm in dictionary){

                ## Look for NAs in the column and replace them with the normal value
                dat[[nm]] <- ifelse(is.na(dat[[nm]]) & 
                                        dat$GENDER == "M", 
                                    normals[normals$variable == val,]$`M`, 
                                    dat[[nm]])
                
                dat[[nm]] <- ifelse(is.na(dat[[nm]]) & 
                                        dat$GENDER == "F",
                                    normals[normals$variable == val, ]$`F`, 
                                    dat[[nm]])
        }
    }
    ## Correct heights * weights
    dat[(is.na(dat$HEIGHT_AVG) & dat$GENDER == "M"), ]$HEIGHT_AVG <- normals[normals$variable == "HEIGHT_", ]$M
    dat[(is.na(dat$HEIGHT_AVG) & dat$GENDER == "F"), ]$HEIGHT_AVG <- normals[normals$variable == "HEIGHT_", ]$`F`
    ## Return
    return(dat)
}
```

#### median_replace

Look for `NA` values, replace with median value

```{r}
median_replace <- function(dat){
  for (var in colnames(dat)){
      if (is.numeric(dat[[var]])){
        #print(var)
        med_val <- median(dat[[var]], na.rm = T)
        #print(med_val)
        dat[[var]] <- ifelse(is.na(dat[[var]]), med_val, dat[[var]])
      }
        
  }
  return(dat)
}
```

## Load Data

```{r}
dat <- read.csv("~/vent_trach_study/data/final_trach_cohort.csv", header = T, stringsAsFactors = F)

## Subset data to remove unnecessary variables
dat <- dat[, c("SUBJECT_ID",
               "HADM_ID",
               "ICUSTAY_ID",
               "ADMISSION_AGE",
               "GENDER",
               "ETHNICITY",
               "ADMISSION_TYPE",
               "ADMISSION_LOCATION",
               "FIRST_CAREUNIT",
               "MORTALITY_3MO_DISCH",
               "MORTALITY_1YR_DISCH",
               "COHORT")]

## Remove all duplicates
dat <- dat[!duplicated(dat),]

## Check dimensions
## dim(dat)

## Look
## head(dat)
```

### Vitals

__Necessary Queries:__

1. vitals-first-day.sql
2. vitals-first-vent-day.sql
3. vitals-seventh-vent-day.sql

**min, max, mean** of:

1. Heartrate
2. Systolic Blood Pressure
3. Diastolic Blood Pressure
4. Respiratory Rate
5. Temperature
6. Oxygen Saturation
7. Blood Glucose

```{r}
## Vitals Day 1 (ICU Stay)
vd1 <- read.csv("~/vent_trach_study/data/tables/vitals-first-day.csv", header = T, stringsAsFactors = F)
colnames(vd1) <- toupper(colnames(vd1))
vd1 <- rename(vd1, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "VD1_")

## Vitals Mechanical Ventilation Day 1
vmvd1 <- read.csv("~/vent_trach_study/data/tables/vitals-first-vent-day.csv", header = T, stringsAsFactors = F)
colnames(vmvd1) <- toupper(colnames(vmvd1))
vmvd1 <- rename(vmvd1, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "VMVD1_")

## Vitals Mechanical Ventilation Day 7
vmvd7 <- read.csv("~/vent_trach_study/data/tables/vitals-seventh-vent-day.csv", header = T, stringsAsFactors = F)
colnames(vmvd7) <- toupper(colnames(vmvd7))
vmvd7 <- rename(vmvd7, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "VMVD7_")

## Left join vd1, vdmvd1, 7
tmp <- merge(x = dat, y = vd1, by =c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
tmp <- merge(x = tmp, y = vmvd1, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
tmp <- merge(x = tmp, y = vmvd7, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)

## Clean
rm(vd1, vmvd1, vmvd7)
dat <- tmp
rm(tmp)
```

### Labs

__Necessary Queries:__

1. labs-first-day.sql
2. labs-first-vent-day.sql
3. labs-seventh-vent-day.sql

**min, max, mean** of:

1. Anion Gap
2. Albumin
3. Bands
4. Bicarbonate
5. Bilirubin
6. Creatinine
7. Chloride
8. Glucose

```{r}
ld1 <- read.csv("~/vent_trach_study/data/tables/labs-first-day.csv", header = T, stringsAsFactors = F)
colnames(ld1) <- toupper(colnames(ld1))
ld1 <- rename(ld1, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "CHARTTIME"), "LD1_")

lmvd1 <- read.csv("~/vent_trach_study/data/tables/labs-first-vent-day.csv", header = T, stringsAsFactors = F)
colnames(lmvd1) <- toupper(colnames(lmvd1))
lmvd1 <- rename(lmvd1, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "CHARTTIME"), "LMVD1_")

lmvd7 <- read.csv("~/vent_trach_study/data/tables/labs-seventh-vent-day.csv", header = T, stringsAsFactors = F)
colnames(lmvd7) <- toupper(colnames(lmvd7))
lmvd7 <- rename(lmvd7, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "CHARTTIME"), "LMVD7_")

## Left join vd1, vdmvd1, 7
tmp <- merge(x = dat, y = ld1, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
tmp <- merge(x = tmp, y = lmvd1, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
tmp <- merge(x = tmp, y = lmvd7, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)

## Clean
rm(ld1, lmvd1, lmvd7)
dat <- tmp
rm(tmp)
```

### Height and Weight Data

__Necessary Queries:__

1. heightfirstday.sql
2. weight-first-day.sql
3. weight-first-vent-day.sql
4. weight-seventh-vent-day.sql

```{r}
## Height is constant
height <- read.csv("~/vent_trach_study/data/tables/heightfirstday.csv", header = T, stringsAsFactors = F)
colnames(height) <- toupper(colnames(height))

## Find mean height from all sources
height$HEIGHT_AVG <- apply(height[,2:4], 1, mean, na.rm = T)

## Weight can be pulled for different days
wd1 <- read.csv("~/vent_trach_study/data/tables/weightfirstday.csv", header = T, stringsAsFactors = F)
colnames(wd1) <- toupper(colnames(wd1))
## Take the average weight
wd1$WEIGHT_AVG <- apply(wd1[,2:5], 1, mean, na.rm = T)
wd1 <- aggregate(WEIGHT_AVG ~ ICUSTAY_ID, data = wd1, FUN = mean)
wd1 <- rename(wd1, "ICUSTAY_ID", "WEIGHTD1_")

wmvd1 <- read.csv("~/vent_trach_study/data/tables/weight-first-vent-day.csv", header = T, stringsAsFactors = F)
colnames(wmvd1) <- toupper(colnames(wmvd1))
## Take the average weight
wmvd1$WEIGHT_AVG <- apply(wmvd1[,2:5], 1, mean, na.rm = T)
wmvd1 <- aggregate(WEIGHT_AVG ~ ICUSTAY_ID, data = wmvd1, FUN = mean)
wmvd1 <- rename(wmvd1, "ICUSTAY_ID", "WEIGHTMVD1_")

wmvd7 <- read.csv("~/vent_trach_study/data/tables/weight-seventh-vent-day.csv", header = T, stringsAsFactors = F)
colnames(wmvd7) <- toupper(colnames(wmvd7))
## Take the average weight
wmvd7$WEIGHT_AVG <- apply(wmvd7[,2:5], 1, mean, na.rm = T)
wmvd7 <- aggregate(WEIGHT_AVG ~ ICUSTAY_ID, data = wmvd7, FUN = mean)
wmvd7 <- rename(wmvd7, "ICUSTAY_ID", "WEIGHTMVD7_")

## Left join vd1, vdmvd1, 7
tmp <- merge(x = dat, y = height, by = "ICUSTAY_ID")

## Keep only average
tmp$HEIGHT <- NULL
tmp$HEIGHT_CHART <- NULL
tmp$HEIGHT_ECHO <- NULL

tmp <- merge(x = tmp, y = wd1, by = c("ICUSTAY_ID"), all.x = TRUE)
tmp <- merge(x = tmp, y = wmvd1, by = c("ICUSTAY_ID"), all.x = TRUE)
tmp <- merge(x = tmp, y = wmvd7, by = c("ICUSTAY_ID"), all.x = TRUE)

## Clean
rm(height, wd1, wmvd1, wmvd7)
dat <- tmp
rm(tmp)
```

### Renal Replacement Therapy

__Necessary Queries:__

1. rrt-first-day.sql
2. rrt-first-vent-day.sql
3. rrt-seventh-vent-day.sql

```{r}
rrtd1 <- read.csv("~/vent_trach_study/data/tables/rrt-first-day.csv", header = T, stringsAsFactors = F)
colnames(rrtd1) <- toupper(colnames(rrtd1))
rrtd1 <- aggregate(RRT ~ ICUSTAY_ID, data = rrtd1, FUN = max)
rrtd1 <- rename(rrtd1, c("ICUSTAY_ID"), "RRTD1_")

rrtmvd1 <- read.csv("~/vent_trach_study/data/tables/rrt-first-vent-day.csv", header = T, stringsAsFactors = F)
colnames(rrtmvd1) <- toupper(colnames(rrtmvd1))
rrtmvd1 <- aggregate(RRT ~ ICUSTAY_ID, data = rrtmvd1, FUN = max)
rrtmvd1 <- rename(rrtmvd1, c("ICUSTAY_ID"), "RRTMVD1_")

rrtmvd7 <- read.csv("~/vent_trach_study/data/tables/rrt-seventh-vent-day.csv", header = T, stringsAsFactors = F)
colnames(rrtmvd7) <- toupper(colnames(rrtmvd7))
rrtmvd7 <- aggregate(RRT ~ ICUSTAY_ID, data = rrtmvd7, FUN = max)
rrtmvd7 <- rename(rrtmvd7, c("ICUSTAY_ID"), "RRTMVD7_")

## Left join rrtd1, rrtmvd1, 7
tmp <- merge(x = dat, y = rrtd1, by = "ICUSTAY_ID", all.x = TRUE)
tmp <- merge(x = tmp, y = rrtmvd1, by = "ICUSTAY_ID", all.x = TRUE)
tmp <- merge(x = tmp, y = rrtmvd7, by = "ICUSTAY_ID", all.x = TRUE)

## Clean
rm(rrtd1, rrtmvd1, rrtmvd7)
dat <- tmp
rm(tmp)
```

## Vasopressors

```{r}
## Reload entire cohort and join to vasopressor durations for time
mvc <- read.csv("~/vent_trach_study/data/final_trach_cohort.csv", header = T, stringsAsFactors = F)

vaso <- read.csv("~/vent_trach_study/data/tables/vasopressordurations.csv", header = T, stringsAsFactors = F)
colnames(vaso) <- toupper(colnames(vaso))

## Temporary merge for calculations
tmp <- merge(mvc[,c("ICUSTAY_ID", "FIRST_VENT_STARTTIME", "INTIME")], vaso, by = "ICUSTAY_ID", all.x = TRUE)

## Use properties of vasopressor durations to check vasopressor use.
## Again, add 0.25day buffer
tmp$VPD1_VP <- ifelse((((as.POSIXlt(as.character(tmp$ENDTIME), tz = "EST")) + (0.25 * (24 * 60 * 60))) >= (as.POSIXlt(as.character(tmp$INTIME), tz = "EST") + (1 * (24 * 60 * 60) ))), 1, 0)
tmp$VPMVD1_VP <- ifelse((((as.POSIXlt(as.character(tmp$ENDTIME), tz = "EST")) + (0.25 * (24 * 60 * 60))) >= (as.POSIXlt(as.character(tmp$FIRST_VENT_STARTTIME), tz = "EST") + (1 * (24 * 60 * 60) ))), 1, 0)
tmp$VPMVD7_VP <- ifelse((((as.POSIXlt(as.character(tmp$ENDTIME), tz = "EST")) + (0.25 * (24 * 60 * 60))) >= (as.POSIXlt(as.character(tmp$FIRST_VENT_STARTTIME), tz = "EST") + (7 * (24 * 60 * 60) ))), 1, 0)

## Correct NAs to 0
tmp$VPD1_VP <- ifelse(is.na(tmp$VPD1_VP), 0, tmp$VPD1_VP)
tmp$VPMVD1_VP <- ifelse(is.na(tmp$VPMVD1_VP), 0, tmp$VPMVD1_VP)
tmp$VPMVD7_VP <- ifelse(is.na(tmp$VPMVD7_VP), 0, tmp$VPMVD7_VP)

tmp <- aggregate(cbind(VPD1_VP, VPMVD1_VP, VPMVD7_VP) ~ ICUSTAY_ID, data = tmp, FUN = max)

## Merge back to data
dat <- merge(x = dat, y = tmp, by = c("ICUSTAY_ID"), all.x = TRUE)

## Clean
rm(mvc, vaso, tmp)
```

## Missing Value Normal Range Correction

```{r}
tmp <- read.csv("~/vent_trach_study/data/normal_reference30Jul19.csv", header = T, stringsAsFactors = F)

## Capture anything coming before a '_' (ALBUMIN, HEMOGLOBIN, etc.)
## Replace captured group including '_' because of PT vs PTT
tmp$variable <- gsub("([^_]*).*", "\\1_", tmp$variable)

## Remove resulting duplicates
tmp <- tmp[!duplicated(tmp),]

## Drop characters from values
tmp$value_m <- as.numeric(unlist(regmatches(tmp$value_m, gregexpr("[-+]?[0-9]*\\.?[0-9]+([eE][-+]?[0-9]+)?", tmp$value_m))))
tmp$value_f <- as.numeric(unlist(regmatches(tmp$value_f, gregexpr("[-+]?[0-9]*\\.?[0-9]+([eE][-+]?[0-9]+)?", tmp$value_f))))

## Correct column names
colnames(tmp) <- c("variable", "M", "F")

## Load normal ranges
nm_range <- read.csv("~/vent_trach_study/data/normal_range30Jul19.csv", header = T, stringsAsFactors = F)

## NA values that are out of range
dat <- outlier_na(dat, nm_range)

## Replace NAs if a replacement is available
dat <- outlier_replace(dat, tmp)

## If no replacement is available, replace with median value
dat <- median_replace(dat)

## Clean environment
rm(tmp, nm_range)
```

## BMI Calculation

```{r}
dat$BMID1 <- dat$WEIGHTD1_WEIGHT_AVG/dat$HEIGHT_AVG/dat$HEIGHT_AVG*10000
dat$BMIMVD1 <- dat$WEIGHTMVD1_WEIGHT_AVG/dat$HEIGHT_AVG/dat$HEIGHT_AVG*10000
dat$BMIMVD7 <- dat$WEIGHTMVD7_WEIGHT_AVG/dat$HEIGHT_AVG/dat$HEIGHT_AVG*10000
```

## Comorbidities on Hospital Admission

```{r}
elixhauser <- read.csv("~/vent_trach_study/data/tables/elixhauser.csv", header = T, stringsAsFactors = F)
colnames(elixhauser) <- toupper(colnames(elixhauser))
elixhauser <- rename(elixhauser, "HADM_ID", "HOSP_ADMIT_")

## Merge
dat <- merge(x = dat, y = elixhauser, by = "HADM_ID", all.x = TRUE)

## Clean
rm(elixhauser)
```

## Illness Severity Scores (ICU Admission Level, day one)

These are all calculated from first ICU stay

```{r, eval = F}
aps <- read.csv("~/vent_trach_study/data/tables/apsiii.csv", header = T, stringsAsFactors = F)
colnames(aps) <- toupper(colnames(aps))

## Subset only necesary data
aps <- aps[,c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "APSIII")]
aps <- rename(aps, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "ICU_D1_")
## Merge
dat <- merge(x = dat, y = aps, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
rm(aps)

oasis <- read.csv("~/vent_trach_study/data/tables/oasis.csv", header = T, stringsAsFactors = F)
colnames(oasis) <- toupper(colnames(oasis))
## Subset
oasis <- oasis[, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "OASIS")]
oasis <- rename(oasis, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "ICU_D1_")
## Merge
dat <- merge(x = dat, y = oasis, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
rm(oasis)

sofa <- read.csv("~/vent_trach_study/data/tables/sofa.csv", header = T, stringsAsFactors = F)
colnames(sofa) <- toupper(colnames(sofa))
## Subset
sofa <- sofa[, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "SOFA")]
sofa <- rename(sofa, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "ICU_D1_")
## Merge
dat <- merge(x = dat, y = sofa, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
rm(sofa)

lods <- read.csv("~/vent_trach_study/data/tables/lods.csv", header = T, stringsAsFactors = F)
colnames(lods) <- toupper(colnames(lods))
## Subset
lods <- lods[, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "LODS")]
lods <- rename(lods, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "ICU_D1_")
## Merge
dat <- merge(x = dat, y = lods, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
rm(lods)

saps <- read.csv("~/vent_trach_study/data/tables/sapsii.csv", header = T, stringsAsFactors = F)
colnames(saps) <- toupper(colnames(saps))
## Subset
saps <- saps[, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID", "SAPSII")]
saps <- rename(saps, c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), "ICU_D1_")
## Merge
dat <- merge(x = dat, y = saps, by = c("SUBJECT_ID", "HADM_ID", "ICUSTAY_ID"), all.x = TRUE)
rm(saps)
```

## Write

```{r}
write.csv(dat, "~/vent_trach_study/data/vent_trach_cohort_clean25May2020.csv", row.names = F)
```
